---
title: "Calibration workflow"
author: "Zacharias Steinmetz"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
    fig_width: 8
vignette: >
  %\VignetteIndexEntry{Calibration workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "# >"
)

data.table::setDTthreads(2)
```

## Packages

For the complete workflow, **data.table** and **ggplot2** are required besides
**envalysis**.

```{r packages}
library(envalysis)
library(data.table)
library(ggplot2)
```

## Sample data

The sample data used stems from Steinmetz et al. (2019). It consists of two
tables: a sequence table and a sample table.

The sequence table contains gas-chromatography/mass spectrometry measurement
data of two phenolic compounds, these are tyrosol and vanillin. Besides the
samples, standard mixtures and extraction blanks (type) were acquired in three
separate analysis batches. Each measurement resulted in an integrated peak area.

```{r, echo=F}
knitr::kable(phenolics$seq[c(1:4,73:74,76,78,80,83:84),], "simple",
             row.names = F)
```

The sample table describes the samples' origin from a 29-day degradation
experiment, in which the phenolic compounds were either degraded in the dark by
the native soil microbial community or photooxidized under UV irradiation after
sterilizing the soil. The samples were processed in threefold replication. Their
weight [g], the volume [mL] of extract solution, and the dilution factor were
recorded.

```{r, echo=F}
knitr::kable(phenolics$samples[c(1:2,4,41:42),], "simple",
             row.names = F)
```

In **envalysis**, the sample data is stored in a two-item list called
`phenolics`. The list items are named `seq` and `samples`.

```{r data}
data("phenolics")
str(phenolics)
```

## Simple calibation

Since the two phenolic compounds were analyzed in three different batches, six
individual calibration curves are required for quantification. For better
understanding, the calibration workflow is first shown for a subset of data,
namely the first batch of tyrosol measurements. The subset is stored in
`tyrosol_1`.

All standards in the `tyrosol_1` subset are used for calibration. The
`'calibration'` object is stored as `cal_1`, which can be printed for additional
information including limits of detection and quantification, the adjusted
*R*^2^, blanks, and statistical checks of the underlying calibration model.

```{r simple_calibration, fig.align="center"}
tyrosol_1 <- subset(phenolics$seq, Compound == "Tyrosol" & Batch == 1)

cal_1 <- calibration(Area ~ `Spec Conc`,
                     data = subset(tyrosol_1, Type == "Standard"))

print(cal_1)
plot(cal_1)
```

Based on `cal_1`, the tyrosol concentrations can be calculated for all samples
using `inv_predict()`. The argument `below_lod = 0` specifies that
concentrations below limit of detection (LOD) should be set to zero.

```{r simple_inv_predict}
tyrosol_1$`Calc Conc` <- inv_predict(cal_1, tyrosol_1$Area, below_lod = 0)
head(tyrosol_1)
```

## Working with `data.table`s

To process all compounds and analysis batches together, the `phenolics` data is
converted to `data.table`s.

```{r data.table}
dt <- lapply(phenolics, as.data.table)
```

To replicate the following steps, try to organize your data in the same way as
shown before. If you want to read in your data directly as `data.table`, use
their `fread()` function, for instance.

## Batch calibration

Subsequently, `calibration()` and `inv_predict()` are applied by compound and
batch.

```{r calibration}
dt$seq[, `Calc Conc` := calibration(Area ~ `Spec Conc`, .SD[Type == "Standard"]) |> 
         inv_predict(Area, below_lod = 0),
       by = .(Compound, Batch)]
head(dt$seq)
```

Calibration parameters like LODs, LOQs, or adjusted *R*^2^ may be stored in a
separate list item for later use.

```{r parameters}
dt$cal <- dt$seq[Type == "Standard", calibration(Area ~ `Spec Conc`) |> 
                   as.list(c("coef", "adj.r.squared", "lod", "loq")),
                 by = .(Compound, Batch)]
print(dt$cal)
```

## Blank subtraction

With the calculated concentrations at hand, the sample concentrations are
subtracted by the extraction blanks to correct for potential lab-borne
contamination.

```{r blank_subtr}
dt$seq[, `Clean Conc` := `Calc Conc` - mean(
  `Calc Conc`[Type == "Extraction blank"], na.rm = T),
  by = .(Batch, Compound)]
```

## Merging tables

The sequence table is merged with the sample table and the contents of
phenolic compounds are calculated from the extraction volume, sample weight, and
dilution factor.

```{r merging}
dt$res <- merge(dt$seq, dt$samples, by = "Name")

dt$res[, Content := `Clean Conc` * (Extract / Weight) * Dilution]
head(dt$res)
```

## Plotting

For plotting the data using **ggplot2**, the contents are summarized by mean and
confidence interval (CI).

```{r plotting, fig.align="center", fig.height=3.5}
dt$sum <- dt$res[, .(Content = mean(Content, na.rm = T),
                     CI = CI(Content, na.rm = T)),
                 by = .(Compound, Treatment, Day)]

ggplot(dt$sum, aes(x = Day, y = Content)) +
  geom_errorbar(aes(ymin = Content - CI, ymax = Content + CI, group = Treatment),
                width = 1, position = position_dodge(1)) +
  geom_point(aes(shape = Treatment, fill = Treatment),
             position = position_dodge(1)) +
  xlab("Day of incubation") +
  ylab(expression("Phenolic content"~"["*mg~kg^-1*"]")) +
  facet_wrap(~ Compound, ncol = 2, scales = "free") +
  scale_shape_manual(values = c(21,24)) +
  scale_fill_manual(values = c("black", "white")) +
  theme_publish()
```

## References

Steinmetz, Z., Kurtz, M.P., Zubrod, J.P., Meyer, A.H., Elsner, M., & Schaumann,
G.E. (2019) Biodegradation and photooxidation of phenolic compounds in soil—A
compound-specific stable isotope approach.
*Chemosphere* **230**, 210-218. DOI: [10.1016/j.chemosphere.2019.05.030](https://doi.org/10.1016/j.chemosphere.2019.05.030).
