---
title: "Particle size estimation"
subtitle: > 
  Based on the hydrometer method by ASTM D422-63 (2007) and Bouyoucos (1927)
author: "Zacharias Steinmetz"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
    fig_width: 8
vignette: >
  %\VignetteIndexEntry{Particle size estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "# >"
)
```

## Materials

### Equipment

* Balance
* Horizontal shaker
* 151H or 152H hydrometer
* 1 L volumetric flask
* 1 L volumetric cylinders for sedimentation
* 500 mL jars with a screw cap
* Squirt bottle
* Thermometer
* Parafilm

### Chemicals

* Sodium hexametaphosphate (SHMP)
* Deionized water conditioned at ambient temperature

### Specimen

* Soil samples, dried, and sieved to 2 mm

## Method

### Dispersing solution

Dissolve 40 g of SHMP in 1 L of deionized water. To facilitate complete
dissolution, fill a 1 L volumetric flask with about 500 mL of deionized water
and add the SHMP step-by-step while stirring the solution with a magnetic
stirrer. The solution is stable for at least one month.

### Calibration

Hydrometers are graduated to be read at the bottom of the meniscus. However,
readings may need be taken at the top of the meniscus in opaque soil
suspensions.

Lower the hydrometer into a 1 L cylinder filled with 125 mL of SHMP solution and
875 mL of deionized water. When stabilized (after about 20 s), record both the
reading at the bottom (zero or composite correction factor) and at the top of
the meniscus (meniscus correction factor).

### Sample pretreatment

If the soil sample contains primarily clay or silt, weigh out 50 g of soil
(accuracy ± 0.01 g) in a 500 mL screw bottle (in 3-fold replication). If the
sample is sandy, take 100 g each. Add 125 mL SHMP solution to the soil. Agitate
the mixture for 16 h (or overnight) in a horizontal shaker.

Completely transfer the soil slurry into the 1 L sedimentation cylinder using a
squirt bottle. Fill up to the 1 L mark with deionized water. Cap the cylinder
with parafilm (ensure tightness). Agitate the slurry by turning the cylinder
upside down and back 30 times during 1 min.

### Measurement

Place the sedimentation cylinder at a bench where it can stand
undisturbed for the next 4–6 h.

Insert the hydrometer into the soil solution as soon as possible and
take the first reading the latest after 40 s.
Measure the temperature of the slurry. Repeat this measurement at 2, 5,
15, 30, 60, 120, 240, and 360 min. In any case, record the exact reading
time. The measurements may be reduced to an initial reading after 40 s
and a second one after 240 or 360 min (Ashworth et al., 2001).
Remove the hydrometer after each measurement, in order not to disturb the 
sedimentation process, and rinse the hydrometer with deionized water.

If the ambient temperature changes, recalibrate the hydrometer.
In case foam obscures the hydrometer reading, use another point on the
hydrometer scale as reference and mark its position with a permanent marker on
the outside of the glass cylinder. The exact readings may be recorded at the
end of the measurement by reinserting the hydrometer at the previous position.

### Data recording

The following table may serve as a sample template to record the data

| Sample data | Time [min] | Hydrometer reading | Correction | Temperature [°C] |
| ----------- | ---------- | ------------------ | ---------- | ---------------- |
| S1          |  0.66      | 39                 | 2          | 23               |
| S1          |  2.00      | 33                 | 2          | 23               |
| S1          |  5.00      | 29                 | 2          | 23               |
| S1          | 15.00      | 23                 | 2          | 23               |
| S1          | 30.00      | 22                 | 2          | 23               |
| S1          | 60.00      | 20                 | 2          | 23               |
| S1          | ...        | ...                | ...        | ...              |
| S2          | ...        | ...                | ...        | ...              |

### Data evaluation

A complete description of how to calculate the particle size distribution from
the recorded hydrometer readings is available in the respective ASTM guideline
(ASTM D422-63, 2007). The algorithm has also been implemented into this package
using the `texture()` function.

After loading **envalysis**, the `texture()` function is applied to the
`clayloam` sample data set coming with this package.

```{r texture}
library(envalysis)

data(clayloam)
clayloam

tex <- texture(reading ~ blank + time + temperature, data = clayloam)

print(tex)
plot(tex)
```

Further soil classification and plotting may be performed using the
**soiltexture** package.

```{r soiltexture_package}
library(soiltexture)
```

For that, the soil texture data needs to be converted into a special
`data.frame`.

```{r soiltexture_convert}
germansoil <- as_tridata(tex, which = "din")
ussoil <- as_tridata(tex, which = "usda")
```

Now, texture classes are determined, for example, in accordance with the German
"Bodenartendiagramm" (DE.BK94.TT) or USDA (USDA.TT).

```{r soiltexture_classify}
TT.points.in.classes(germansoil, class.sys = "DE.BK94.TT")
TT.points.in.classes(ussoil, class.sys = "USDA.TT")
```

The analyzed soil is a clay loam (German: "Toniger Lehm", Lt2), which may be
plotted as follows.

```{r soiltexture_plot, echo = -1}
par(cex = 0.75, cex.lab = 0.5, cex.axis = 0.5, cex.main = 1)
TT.plot(class.sys = "DE.BK94.TT", tri.data = germansoil)
```

For further details, see the **soiltexture** package vignette on
[CRAN](https://cran.r-project.org/package=soiltexture).

## References

Ashworth, J., Keyes, D., Kirk, R., & Lessard, R. (2001). Standard Procedure in
the Hydrometer Method for Particle Size Analysis. *Communications in Soil
Science and Plant Analysis* **32**, 633-642. DOI:
[10.1081/CSS-100103897](https://doi.org/10.1081/CSS-100103897).

ASTM D422-63 (2007). *Standard Test Method for Particle-Size Analysis of Soils*.
Technical standard. ASTM International, West Conshohocken, PA.

Bouyoucos, G.J., (1927). The hydrometer as a new method for the mechanical
analysis of soils. Soil Science 23, 343–354.

Moeys, J., Shangguan, W., Petzold, R., Minasny, B., Rosca, B., Jelinski, N.,
Zelazny, W., Souza, R.M.S., Safanelli, J.L., & ten Caten, A. (2018).
*soiltexture: Functions for Soil Texture Plot, Classification and
Transformation*. URL:
[https://CRAN.R-project.org/package=soiltexture](https://CRAN.R-project.org/package=soiltexture).
