---
title: "Particle size estimation using the hydrometer method modified from ASTM D422-63 (2007) and Bouyoucos (1927)"
author: "Zacharias Steinmetz"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Particle size estimation}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE, 
  message = FALSE,
  comment = "# >"
)
```

## Materials

### Equipment

* Balance
* Horizontal shaker
* 151H or 152H hydrometer
* 1 L volumetric flask
* 1 L volumetric cylinders for sedimentation
* 500 mL jars with a screw cap
* Squirt bottle
* Thermometer
* Parafilm

### Chemicals

* Sodium hexametaphosphate (SHMP)
* Deionized water conditioned at ambient temperature

### Specimen

* Soil samples, dried, and sieved to 2 mm

## Method

### Dispersing solution

Dissolve 40 g of SHMP in 1 L of deionized water.
To facilitate complete dissolution, fill a 1 L volumetric flask with about
500 mL of deionized water and add the SHMP step-by-step
while stirring the solution with a magnetic stirrer.
The solution is stable for at least one month.

### Calibration

Hydrometers are graduated to be read at the bottom of the meniscus.
However, readings may need be taken at the top of the meniscus in opaque
soil suspensions.

Lower the hydrometer into a 1 L cylinder filled with 125 mL of SHMP solution
and 875 mL of deionized water.
When stabilized (after about 20 s), record both the reading at the bottom (zero
or composite correction factor) and at the top of the meniscus
(meniscus correction factor).

### Sample pretreatment

If the soil sample contains primarily clay or silt, weigh out 50 g of
soil (accuracy ± 0.01 g) in a 500 mL screw bottle (in 3-fold
replication). If the sample is sandy, take 100 g each. Add 125 mL SHMP solution
to the soil. Agitate the mixture for 16 h (or overnight) in a horizontal shaker.

Completely transfer the soil slurry into the 1 L sedimentation cylinder
using a squirt bottle. Fill up to the 1 L mark with deionized water.
Cap the cylinder with parafilm (ensure tightness).
Agitate the slurry by turning the cylinder upside down and back 30 times
during 1 min.

### Measurement

Place the sedimentation cylinder at a bench where it can stand
undisturbed for the next 4–6 h.

Insert the hydrometer into the soil solution as soon as possible and
take the first reading the latest after 40 s.
Measure the temperature of the slurry. Repeat this measurement at 2, 5,
15, 30, 60, 120, 240, and 360 min. In any case, record the exact reading
time. The measurements may be reduced to an initial reading after 40 s
and a second one after 240 or 360 min (Ashworth et al., 2001).
Remove the hydrometer after each measurement, in order not to disturb the 
sedimentation process, and rinse the hydrometer with deionized water.

If the ambient temperature changes, recalibrate the hydrometer.
In case foam obscures the hydrometer reading, use another point on the
hydrometer scale as reference and mark its position with a permanent marker on
the outside of the glass cylinder. The exact readings may be recorded at the
end of the measurement by reinserting the hydrometer at the previous position.

### Data recording

The following table may serve as a sample template to record the data

| Sample data | Time [min] | Hydrometer reading | Correction | Temperature [°C] |
| ----------- | ---------- | ------------------ | ---------- | ---------------- |
| S1          |  0.66      | 39                 | 2          | 23               |
| S1          |  2.00      | 33                 | 2          | 23               |
| S1          |  5.00      | 29                 | 2          | 23               |
| S1          | 15.00      | 23                 | 2          | 23               |
| S1          | 30.00      | 22                 | 2          | 23               |
| S1          | 60.00      | 20                 | 2          | 23               |
| S1          | ...        | ...                | ...        | ...              |
| S2          | ...        | ...                | ...        | ...              |

### Data evaluation

A complete description of how to calculate the particle size
distribution from the recorded hydrometer readings is available in the
respective ASTM guideline (ASTM D422-63, 2007). The
algorithm has also been implemented into this package using the `texture()`
function:

```{r texture}
# Load envalysis
library(envalysis)

# Load and look at sample data
data(clayloam)
clayloam

# Calculate the particle size distribution
tex <- texture(reading ~ blank + time + temperature, clayloam, plot = T)
tex
```

Further soil classification and plotting may be done with the **soiltexture**
package.

```{r classify}
# Load soiltexture
library(soiltexture)

# Prepare data
germansoil <- data.frame(t(tex$din["Estimate",] * 100))
names(germansoil) <- toupper(names(germansoil))

ussoil <- data.frame(t(tex$usda["Estimate",] * 100))
names(ussoil) <- toupper(names(ussoil))

# Get texture class, for example, in accordance with the German
# "Bodenartendiagramm" (DE.BK94.TT)
TT.points.in.classes(germansoil, class.sys = "DE.BK94.TT")

# Get USDA texture class (USDA.TT)
TT.points.in.classes(ussoil, class.sys = "USDA.TT")
```

The analyzed soil is a clay loam (German: "Toniger Lehm", Lt2).

## References

Ashworth, J., Keyes, D., Kirk, R., & Lessard, R. (2001). Standard Procedure in the 
Hydrometer Method for Particle Size Analysis. *Communications in Soil Science and 
Plant Analysis* **32**, 633-642. DOI: [10.1081/CSS-100103897](https://doi.org/10.1081/CSS-100103897).

ASTM D422-63 (2007). *Standard Test Method for Particle-Size Analysis of 
Soils*. Technical standard. ASTM International, West Conshohocken, PA.

Bouyoucos, G.J., (1927). The hydrometer as a new method for the mechanical 
analysis of soils. Soil Science 23, 343–354.

Moeys, J., Shangguan, W., Petzold, R., Minasny, B., Rosca, B., Jelinski, N.,
Zelazny, W., Souza, R.M.S., Safanelli, J.L., & ten Caten, A. (2018). *soiltexture:
Functions for Soil Texture Plot, Classification and Transformation*. URL: [https://CRAN.R-project.org/package=soiltexture](https://CRAN.R-project.org/package=soiltexture).
